---
title: 'Practical Machine Learning Project'
author: 'MartinSRC, June, 2014'
output:
  html_document:
    toc: true
    number_sections: true
    fig_caption: true
    highlight: tango
  md_document:
    variant: markdown_github
    toc: true
---
*Note.* 'index.html' and 'README.md' were both rendered using [rmarkdown] from the 'index.Rmd' file generated by 'PMLproject.R'. Resulting index.html file placed on github.io (i.e. [Github Pages]) for working html link.
```{r options, echo=F}
library(knitr)
opts_chunk$set(comment='',tidy=F,fig.width=3,fig.height=3,fig.align='center')
```

# Background
The prediction exercise is as follows:
> *"Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset)."*
The goal of the assignment is to predict the provided test set observations.

----

# Data
The data for this project come from [HAR Groupware].  
```{r data}
download.file(
   'http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv',
   destfile='pml-training.csv')
download.file(
   'http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv',
   destfile='pml-testing.csv')
train <- read.csv('pml-training.csv')
test <- read.csv('pml-testing.csv')
```

----

# *"How you built your model"*

## Pre-processing
The first step was data pre-processing. A 70% training and 30% validation set were created because this is a simple and reliable method of validating a machine learner. Predictors with near zero-variance were removed because they do not contribute much to predictive power. Numeric variables were extracted for training to avoid distortion from arbitrary integers. Missing values were treated with imputation to improve accuracy.
```{r preprocessing}
set.seed(432423)
library(caret)

#Data Partition
x <- createDataPartition(train$classe,list=FALSE,p=.7)
modtrain <- train[x,]
validate <- train[-x,]

#Near Zero Variance
blanks <- nearZeroVar(modtrain)
modtrain <- modtrain[-blanks]
validate <- validate[-blanks]
fulltrain <- train[-blanks]
finaltest <- test[-blanks]

#Imputation
numvars <- which(lapply(modtrain,class) %in% c('numeric'))
preObj <- preProcess(modtrain[,numvars],method='knnImpute')
imputed <- predict(preObj,modtrain[,numvars])
modtrain <- cbind(data.frame(classe=modtrain$classe),imputed)
imputed <- predict(preObj,validate[,numvars])
modtest <- cbind(data.frame(classe=validate$classe),imputed)
imputed <- predict(preObj,fulltrain[,numvars])
fulltrain <- cbind(data.frame(classe=fulltrain$classe),imputed)
finaltest <- predict(preObj,finaltest[,numvars])
```

----

## Training
A competitive random forest machine learning algorithm was chosen because:
1. It is competitive in modelling competitions (e.g. kaggle).
2. It can model a categorical response variable.
3. It is a simple yet high performing method for making 'crisp' classifications.
```{r training}
library(randomForest)
modelFit <- randomForest(classe~., modtrain, ntree=500)
```

----

# *"How you used cross validation"*
The prior creation of a cross validation data set allowed assessment od the model both in- and out-of-sample.

## Confirmed In-Sample Error
```{r insampleassessment}
preds <- predict(modelFit, modtrain)
confusionMatrix(preds, modtrain$classe)
```

The high performance (i.e. 100% sample accuracy) is not a reliable estimate of the machine learner. The estimated out-of-sample error will be more reliable.

## Estimated Out-of-Sample Error
```{r outsampleassessment}
preds <- predict(modelFit, modtest)
confusionMatrix(preds, modtest$classe)
```

The accuracy is still very high. This indicates that this will be a high performing model and should perform well on the final 20 test observations. The test set was prepared with imputations above, so the final model was trained on all data and predictions were made for submission.

```{r finalfullmodel}
modelFit <- randomForest(classe~., fulltrain, ntree=500)
preds <- predict(modelFit, modtest)
confusionMatrix(preds, modtest$classe)
answers <- predict(modelFit, finaltest)
answers 
```

----

# Submission
Test files were generated with code then submitted for full credit. It can be concluded that random forest is still a simple an effective machine learner, that is, ideal for 'practical machine learning'.
```{r submission, results='hide'}
n <- length(answers)
sapply(1:n,function(x){write.table(answers[[x]],
   file=paste0('problem_id_',x,'.txt'), eol='',
   quote=F,row.names=F,col.names=F)})
```

----

# References

```{r refs}
library(bibtex)
write.bib(rev(.packages())[-c(2:7)],verbose=F)
read.bib('Rpackages.bib')
```

[HAR Groupware]: http://groupware.les.inf.puc-rio.br/har/ 'HAR Groupware'
[rmarkdown]: http://rmarkdown.rstudio.com/ 'R Markdown: Dynamic Documents for R'
[Github Pages]: https://pages.github.com/ 'Github Pages: Websites for you and your projects'

